Shifting paradigms in Ocean Color: Bayesian Inference for Uncertainty-Aware Chlorophyll Estimation 

### 2025-05-28 manuscript/manuscript.qmd
updated model 6 explanation

---
title: "Shifting paradigms in Ocean Color: Bayesian Inference for Uncertainty-Aware Chlorophyll Estimation "
format:
    agu-pdf:
        keep-tex: true
    agu-html: default
author:  
  - name: Erdem M. Karaköylü
    orcid: 0000-0002-6156-1720
    corresponding: true
    email: erdemk@protonmail.com
    roles:
      - Investigator
      - Software
      - Visualization
      - Project administration
    affiliations:
      - Independant Researcher
  - name: Susanne E. Craig
    orcid: 0000-0002-8963-0951
    corresponding: false
    email: susanne.e.craig@nasa.gov
    roles:
      - Co-Investigator
    affiliations:
      - NASA/UMBC
abstract: Placeholder
plain-language-summary: Placeholder
keywords: ['Bayesian modeling', 'Ocean Color', ]
key-points:
  - Bayesian models have not been fully adopted in Ocean Color remote sensing yet.
  - These models are a principled way to take uncertainty into account
  - 
bibliography: references.bib
citation:
  container-title: Remote Sensing of Environment
keep-tex: true
number-sections: true
date: last-modified
---

@@

This model follows the OC6 formulation proposed by O’Reilly et al. (2019), with the difference that the denominator is the sum of $Rrs555$ and $Rrs670$ rather than the mean. The model mean is a fourth-order polynomial of $log(MBR)$ to estimate $log(Chl)$. The set of proabilistic equations that define this model can be found in the Supplemental Material Section.
This model can be represented by the following set of probabilistic equations where $X$ represents $log(MBR)$, and $y_i$ correspondonds to the likelihood of the $i^{th}$ $log(Chl)$ observation

::: {#fig-oc6-struct}

![](images/oc6_structure.png)

Model 1 DAG showing model parameters and distributions used as priors. The intercept $α$ and the polynomial slopes, $β1$-$β4$ all have Gaussian priors; $σ$, the likelihood dispersion parameter, receives a Gamma prior; the likelihood itself is constrained by a Truncated Normal distribution (see text for more). The plate containing the likelihood is indexed ($obs\_idx$) indicating the likelihood will be computed for each of the 1082 observations contained in the dataframe and show the number of data points available for fitting.

@@

#### Model 2: Hierarchical Linear Regression (4-band Grouping)

*Model 2* introduces partial pooling across four groups defined by the dominant MBR numerator band; Rrs411, Rrs443, Rrs489, Rrs510 (*cf*. @fig-eda-mbr). Each group receives its own slope and intercept with shared hyperpriors. On the one hand each group has its own sub-model that can capture details specific to it, while the shared hyperpriors allow for the sharing of information between groups when fitting the model, thus maximizing the use of information contained in the available data.  Mathematical description of the model can be found in the Supplementary Material. @fig-model2-struct illustrates the model structure.

@@

#### Model 5: Hierarchical Linear Regression with Linear Heteroscedasticity

The data plotted in @fig-eda suggests heteroscedasticity - variance that is not constant, but rather changes depending on the data. Models 3 and 4 investigate the role of group-wise constant heteroscedasticity and global linear heteroscedasticity, respectively. Details of these models can be found in the Supplementary Material section. To investigate this further, Model 5 looks at group-wise linear heteroscedasticity. Thus, for a given MBR group $j$ $σ_j$ is modeled as a log-linear relationship such that $$log(σ) = α_j + β_j × log(MBR)$ and then transformed via $exp(log(σ_j))$ before being passed on to the likelihood. Similar to the model mean,  $μ_j$, $σ_j$'s slope and intercept terms are partially pooled and so have common hyperpriors. Mathematical descriiption of the model can be found in the Supplementary Material section; @fig-model5-struct.  illustrates the model structure.

@@

Model 5 - Multi-level hierarchical linear model with partial pooling across $MBR$  groups (Rrs411, Rrs443, Rrs489, Rrs510) for both model mean and model variance.  Model parameter prior mean and variance have Gaussian and Exponential hyperpriors similar to Model 2s.
:::

####  Model 6:  Measurement-Specific Dispersion

This model adds a measurement-method-specific dispersion term to the structure of Model 6, capturing the known increase in variability associated with fluorescence-based chlorophyll estimates. A reasonable assumption is that HPLC measurements are less noisy than fluorescence. Thus $σ$ is given the expression:

@@

*Priors* - I used weakly informative priors across all models to balance flexibility and regularization.  For regression parameters  I used Gaussian priors, which in the case of single-level models like Model 1, were given a $0$ mean and variance $1$, to encourage parameter sparsity. Hyperpriors of prior means were set to Gaussans of mean 0 and variance 1 to encourage sparsity. Hyperpriors of prior variances set to  Exponential distribution with rate set to $$1$, which has significant density near 0, encourages similarity across groups unless the data mandates otherwise. This hyperprior also encourages uncertainty shrinkage across groups if possible. The likelihood dispersion $σ$, when constant was set, to a Gamma distribution of shape and rate both set to $2$. Prior soundness was assessed using *Prior Predictive Checks*. This procedure takes advantage of the fact taht Bayesian models are generative to insure that values of ($log(Chl)$) predicted by the model are reasonable even before model fitting. This was plotted as Cumulative Distribution Function (CDF) of $log(Chl)$. See *Results* section for a Prior Predictive Checks plot example; plots for all models can be found in the Supplementary Material section.

>>>>

---
title: "Shifting paradigms in Ocean Color: Bayesian Inference for Uncertainty-Aware Chlorophyll Estimation "
format:
    agu-pdf:
        keep-tex: true
    agu-html: default
author:  
  - name: Erdem M. Karaköylü
    orcid: 0000-0002-6156-1720
    corresponding: true
    email: erdemk@protonmail.com
    roles:
      - Investigator
      - Software
      - Visualization
      - Project administration
    affiliations:
      - Independent Researcher
  - name: Susanne E. Craig
    orcid: 0000-0002-8963-0951
    corresponding: false
    email: susanne.e.craig@nasa.gov
    roles:
      - Co-Investigator
    affiliations:
      - NASA/UMBC
abstract: Placeholder
plain-language-summary: Placeholder
keywords: ['Bayesian modeling', 'Ocean Color', ]
key-points:
  - Bayesian models have not been fully adopted in Ocean Color remote sensing yet.
  - These models are a principled way to take uncertainty into account
  - 
bibliography: references.bib
citation:
  container-title: Remote Sensing of Environment
keep-tex: true
number-sections: true
date: last-modified
---

@@

This model follows the OC6 formulation proposed by O’Reilly et al. (2019), with the difference that the denominator is the sum of $Rrs555$ and $Rrs670$ rather than the mean. The model mean is a fourth-order polynomial of $log(MBR)$ to estimate $log(Chl)$. The set of proabilistic equations that define this model can be found in the Supplemental Material Section. The model structure and inference flow is represented in @fig-model1_struct as a Directed Acyclic Graph (DAG)

::: {#fig-oc6-struct pos=H}

![](images/model1_structure.png)

Model 1 Directed Acyclic Graph showing model parameters and their prior distributions and how they relate . The intercept $α$ and the polynomial slopes, $β1$-$β4$ all have Gaussian priors; $σ$, the likelihood dispersion parameter, receives a Gamma prior; the likelihood itself is constrained by a Truncated Normal distribution (see text for more). The plate containing the likelihood is indexed ($obs\_idx$) indicating the likelihood will be computed for each of the 1082 observations contained in the dataframe and show the number of data points available for fitting.

@@

#### Model 2: Hierarchical Linear Regression (HLR)

*Model 2* introduces partial pooling across four groups defined by the dominant MBR numerator band; Rrs411, Rrs443, Rrs489, Rrs510 (*cf*. @fig-eda-mbr). Each group receives its own slope and intercept, and these parameters have shared hyperpriors. Consequently, each group has its own sub-model that can capture details specific to it and the shared hyperpriors allow for the sharing of information between groups when fitting the model. This approach maximizes the use of information contained in the available data. Hierarchical models are usually result in uncertainty shrinkage.  Mathematical description of the model can be found in the Supplementary Material. @fig-model2-struct illustrates the model structure.

@@

#### Model 5: HLR with Added Linear Heteroscedastic Likelihood Dispersion

The data plotted in @fig-eda suggests heteroscedasticity - variance that is not constant, but rather changes depending on the data. Models 3 and 4 investigate the role of group-wise constant heteroscedasticity and global linear heteroscedasticity, respectively. Details of these models can be found in the Supplementary Material section. To investigate this further, Model 5 looks at group-wise linear heteroscedasticity. Thus, for a given MBR group $j$ $σ_j$ is modeled as a log-linear relationship such that $log(σ) = α_j + β_j × log(MBR)$ and then transformed via $exp(log(σ_j))$ before being passed on to the likelihood. Similar to the model mean,  $μ_j$, $σ_j$'s slope and intercept terms are partially pooled and so have common hyperpriors. Mathematical descriiption of the model can be found in the Supplementary Material section; @fig-model5-struct.  illustrates the model structure.

@@

Model 5 - Multi-level hierarchical linear model with partial pooling across $MBR$  groups (Rrs411, Rrs443, Rrs489, Rrs510) for both model mean and model variance.  Model parameter prior mean and variance have Gaussian and Exponential hyperpriors similar to Model 2.
:::

####  Model 6:  Augmenting Model 5 with Measurement-Specific Dispersion

A reasonable assumption is that HPLC measurements are less noisy than fluorescence. This model adds a measurement-method-specific dispersion term to the structure of Model 6, to capture the suspected increase in variability associated with fluorescence-based chlorophyll estimates. A reasonable assumption is that HPLC measurements are less noisy than fluorescence. The likelihood dispersion, $σ$, is given the expression $log(σ) = α_j + β_j × log(MBR) + γ_{chl\_type} * (1 - chl_{type\_idx})$, where $chl_{type\_idx} = 1$ if *in-situ* chlorophyll was measured with HPLC, $0$ otherwise. As a result, $σ$ is the same as in Model 5 if $chl_{type\_idx} = 1$, since the extra term is nullified. In the case of fluorescence measurement, on the other hand, $log(σ)$ receives an extra noise term, providing an effective measurement of noise difference between the two methods.

@fig-model6-struct shows the inference DAG of Model 6.

@@

*Priors* - I used weakly informative priors across all models to balance flexibility with regularization.  For regression parameters  I used Gaussian priors, In the case of the  single-level  Model 1, this meant priors with $0$ mean and variance of $1$ to encourage parameter sparsity. In the case of Models 2-6, which are multi-level models, hyperpriors of prior means were set to Gaussans of mean 0 and variance 1, again to encourage sparsity. Hyperpriors of prior variances were  Exponential distributions with rate set to $1$, which has significant density near 0, encourages similarity across groups unless the data mandates otherwise. This hyperprior also encourages uncertainty shrinkage across groups if possible. In the case of homoscedasticity, or group-wise heteroscedasticity, I used a Gamma prior of shape and scale of $2$ for the likelihood dispersion $σ$, which provides strictly postivie support. Prior soundness was assessed using *Prior Predictive Checks*. This procedure takes advantage of the fact taht Bayesian models are generative, meaning the model can be sampled without data, to insure that values of ($log(Chl)$) predicted by the model are reasonable even before model fitting. Prior predictive $log(Chl)$ values were plotted as Cumulative Distribution Function (CDF) of $log(Chl)$. @fig-model1-priorpc provides

::: {#fig-model1-priorpc fig-pos="H"}

![](images/model1_prior.png)

Prior Predictive Checks, here shown for Model 1. This step allows the practitioner to verify the soundness of model assumptions - choice of priors, the model formulation, etc. This can be done even before data is collected as the generative nature of Bayesian models means output can be produce based solely on the priors and the model's mathematical expression alone. Results are presented as Cumulative Distribution Function. In blue, 500 draws were performed each containing a prescribed number of observations. The dashed line represents the CDF of the simulation mean, In black, the CDF of the data is shown for comparison.

See *Results* section for a Prior Predictive Checks plot example; plots for all models can be found in the Supplementary Material section.
